---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First we load the data.
```{r} 
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

Convert date inputs to variables of the date class.
```{r}
activity$date <- strptime(activity$date, format = "%Y-%m-%d")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```
  
  
## What is mean total number of steps taken per day?  


Calculate the total numbers of steps taken per day, and create a histogram of the values.  
```{r}
dayTotals <- aggregate(steps ~ date, activity, sum)
hist(dayTotals$steps, xlab = "Total Daily Steps", main = "Histogram of Total Daily Steps", breaks = c(0, 2500, 5000, 7500, 10000, 12500, 15000, 17500, 20000, 22500, 25000), ylim = c(0,20))
```

Calculate the mean and median total number of steps taken per day.
```{r}
dayMean <- mean(dayTotals$steps)
dayMean <- formatC(dayMean, digits = 2, format = "f")
dayMedian <- median(dayTotals$steps)
```
  
The mean total daily steps is `r dayMean` and the median total daily steps is `r dayMedian`.


## What is the average daily activity pattern?  

Calculate mean numbers of steps per interval over all days.  Plot calculations in a time series plot.
```{r}
intervalMeans <- aggregate(steps ~ interval, activity, mean)
plot(intervalMeans, type = "l", xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern")
```

Calculate which interval contains the maximum number of steps.
```{r}
maxVal <- max(intervalMeans$steps)
maxSteps <- subset(intervalMeans, intervalMeans$steps == maxVal)
```
The interval `r maxSteps$interval` contains the maximum numbers of steps, at a value of `r maxVal` steps on average.  

## Imputing missing values  

Calculate the total number of NAs present in the dataset.
```{r}
NA.vals <- is.na(activity$steps)
NA.sum <- sum(NA.vals)
```
The total number of NAs present is `r NA.sum`.

We will impute the missing values based on the mean number of steps for that interval.
```{r}
activity_full <- activity
activity_full$steps <- ifelse(is.na(activity_full$steps), intervalMeans$steps[match(activity_full$interval, intervalMeans$interval)], activity_full$steps)
```

We now have a complete data set, with all missing values filled in.  The first 10 rows of the data set are shown below.
```{r}
head(activity_full, 10)
```

A histogram of the total number of steps taken per day, including our imputed values, is shown below.
```{r}
dayTotalsFull <- aggregate(steps ~ date, activity_full, sum)
hist(dayTotalsFull$steps, xlab = "Total Daily Steps", main = "Histogram of Total Daily Steps with Imputed Data", breaks = c(0, 2500, 5000, 7500, 10000, 12500, 15000, 17500, 20000, 22500, 25000), ylim = c(0,30))

dayMeanFull <- mean(dayTotalsFull$steps)
dayMeanFull <- formatC(dayMeanFull, digits = 2, format = "f")
dayMedianFull <- median(dayTotalsFull$steps)
dayMedianFull <- formatC(dayMedianFull, digits = 2, format = "f")
```

The mean of the completed data set is `r dayMeanFull` and the median of the completed data set is `r dayMedianFull`.  
We find that the mean and median of the dataset with the imputed values do not differ wildly from the estimates calculated from the original data set.  The mean has remained the same, and the median has increased slightly.  The histogram of the total daily steps has changed slightly, with an increase in frequency in the daily totals between 10,000 and 12,500 steps.  Overall, it appears that imputing the missing data did not have a large impact on our calculations.

## Are there differences in activity patterns between weekdays and weekends?  


We must create a factor variable to differentiate between weekdays and weekends.This variable will be called 'dayType'.
```{r}
activity_full$dayType <- sapply(activity_full$date, function(x) {
  if (weekdays(x) == "Saturday" | weekdays(x) == "Sunday") 
    {y <- "Weekend"} 
  else {y <- "Weekday"} })
```

We will create a plot to compare the average number of steps taken per interval across all weekdays days and all weekend days.
```{r}
library(ggplot2)
intervalMeansType <- aggregate(steps ~ interval + dayType, activity_full, mean)
g <- ggplot(intervalMeansType, aes(interval, steps, color = dayType))
g <- g + geom_line() + facet_grid(dayType~.) + labs(x = "Interval", y = "Average Number of Steps")
print(g)
```

